How fluid-mechanical erosion creates anisotropic porous media

### 2022-06-28 docs/PhysicsPaper/StokesErosion.tex
wrapped up first read

\author[NJIT]{Shang-Huan Chiu}
%\affiliation{Department of Scientific Computing, Florida State University, Tallahassee, FL, 32306.}

@@

\address[Colgate]{Colgate University}
\address[FSU]{Florida State University}
\address[NJIT]{New Jersey Institute of Technology}

@@

	Flow-induced erosion acts across a range of scales in the natural world, from massive geological structures sculpted by wind or water~\cite{abrams2009growth, perkins2015amplification, mac2020ultra, sharma2022alcove, mac2022morphological}, to mesoscopic patterns formed by surface or internal flows \cite{berhanu2012shape, bertagni2021hydrodynamic, weady2022anomalous}, and down to granular and porous networks slowly disintegrating in groundwater flows \cite{chiu2020viscous, szymczak2009wormhole, jager2017channelization, grodzki2019reactive, bizmark2020multiscale, derr2020flow, zareei2022temporal}. The associated nonlinear feedback between changing shapes and the surrounding flows can imprint across all of these scales, affecting large-scale features as well as small-scale ones, such as the microstructure of porous materials. Porous media encountered in nature, for example, typically exhibit material {\em anisotropy} in that they permit seepage flow in certain directions more easily than in others.  Typical materials are more permeable to flow in the longitudinal (or horizontal) direction than transversely by a factor of 5--20~\cite{bear1988dynamics, anderson2015applied}. Most commonly, this material anisotropy is attributed to the sedimentation process, in which, due to the fluid-structure interaction, non-spherical particles tend to settle with their long axis parallel to the plan of deposition~\cite{bear1988dynamics}. Controlled experiments, however, have not been performed to test this hypothesis, and other mechanisms may be at work. Here, we use highly-accurate numerical simulations to examine an alternative, and possibly complementary, mechanism: namely, that the flow-induced erosion of the medium's solid constituents contributes to its overall anisotropy.

@@

The outline of the paper is as follows. In Section~\ref{sec:formulation} we discuss the governing equations for the fluid flow and for interface evolution. In Section~\ref{sec:DLP} we discuss the numerical methods. In Section~\ref{sec:medium} we measure porous-medium properties, in particular permeability. In Section~\ref{sec:results} we provide results and we conclude in section~\ref{sec:conclusion}.

@@

%-----------------------------------------------------------------------------------------------%
\section{Boundary Integral Equation and Cauchy Integral Formulation}
\label{sec:DLP}
%-----------------------------------------------------------------------------------------------%
To accurately and efficiently solve the Stokes equations~\eqref{eqn:StokesEq}, we reformulate the system as a boundary integral equation (BIE). A BIE formulation has the advantage that all the unknowns are on the one-dimensional boundaries of the domain. That is, only the boundary of the complex geometry must be discretized which we do with a spectrally accurate Fourier basis. Applying the same approach as our previous works~\cite{chiu2020viscous, quaife2018boundary}, we represent the velocity as the sum of a double-layer potential and a combination of Stokeslets and rotlets~\cite{pow-mir1987}
\begin{align}
  \uu(\xx) = \DDD[\eeta](\xx) + \sum_{\ell=1}^{M}\left(
    S[\llambda_\ell](\xx) + R[\xi_\ell](\xx)\right), 
    \quad \xx \in \Omega,
  \label{eqn:velocity}
\end{align}
where 
\begin{align}
  \DDD[\eeta](\xx) = \frac{1}{\pi}\int_{\bd\Omega} 
  \frac{\rr \cdot \nn}{\rho^2} \frac{\rr \otimes \rr}{\rho^2}
  \eeta(\yy) \, ds_\yy,
  \label{eqn:velocityDLP}
\end{align}
where $\rr = \xx - \yy$ and $\rho = \|\rr\|$. Note that $\bd\Omega$ includes both the eroding bodies and the outer boundary. The Stokeslet and the rotlets are
\begin{align}
  S[\lambda_\ell](\xx) = \frac{1}{4\pi} \left(-\log \rho_\ell +
  \frac{\rr_\ell \otimes \rr_\ell}{\rho_\ell^2} \right) \llambda_\ell,
  \quad
  R[\xi_\ell](\xx) = \frac{\rr_\ell^\perp}{\rho_\ell^2}\xi_\ell,
\end{align}
respectively, where $\rr_\ell = \xx - \cc_\ell$, $\rho_\ell = \|\rr_\ell\|$, and $\cc_\ell$ is a point inside body $\ell$. If the density function, Stokeslets, and rotlets satisfy the second-kind boundary integral equation
\begin{subequations}
  \label{eqn:FredBIE}
  \begin{alignat}{3}
    \UU(\xx) &= -\frac{1}{2}\eeta(\xx) + \DDD[\eeta](\xx) + 
      \sum_{\ell=1}^{M}\left(S[\llambda_\ell](\xx) + R[\xi_\ell](\xx)\right), 
      \quad &&\xx \in \Gamma, \\
    \mathbf{0} &= -\frac{1}{2}\eeta(\xx) + \DDD[\eeta](\xx) + 
      \sum_{\ell=1}^{M}\left(S[\llambda_\ell](\xx) + R[\xi_\ell](\xx)\right), 
      &&\xx \in \gamma_\ell, \: \ell = 1,\ldots,M, \\
    \llambda_\ell &= \int_{\gamma_\ell} \eeta(\yy) \, ds_\yy, 
      &&\ell = 1,\ldots,M, \\
    \xi_\ell &= \int_{\gamma_\ell} (\yy - \cc_\ell)^\perp \cdot 
      \eeta(\yy) \, ds_\yy, &&\ell = 1,\ldots,M,
  \end{alignat}
\end{subequations}
then the representation~\eqref{eqn:velocity} satisfies the Stokes equations with the required boundary conditions~\eqref{eqn:StokesEq}. We solve~\eqref{eqn:FredBIE} by discretizing $\gamma_\ell$ and $\Gamma$ at equispaced collocation points, and then replacing the integrals with quadrature rules. This results in a linear system with a mesh-independent condition number, and it is solved iteratively with GMRES.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Quadrature for Cauchy integrals}
The accuracy of our method is determined by the accuracy of the
quadrature rule applied to equation~\eqref{eqn:FredBIE}. Since we have
written the Stokes double-layer potential
velocity~\eqref{eqn:velocityCauchy}, deformation
tensor~\eqref{eqn:deformationCauchy}, and
vorticity~\eqref{eqn:vorticityCauchy} as a sum of Cauchy integrals and
their derivatives, the accuracy of the method depends on the accuracy of
computing a general Cauchy integral
\begin{align}
  v(x) = \frac{1}{2\pi i} \int_{\gamma} \frac{\eta(y)}{x-y} \, dy.
  \label{eqn:cauchy1}
\end{align}
Here we describe a quadrature formulae that was first used to
approximate analytic functions~\cite{ioa-pap-per1991}, and then extended
to Stokes layer potentials~\cite{bar-wu-vee2015}. The quadrature method
requires the boundary data of the analytic function~\eqref{eqn:cauchy1}
which satisfies the Sokhotski-Plemelj jump relation
\begin{align}
  v^{-}(x_0) = \lim_{\substack{x \rightarrow x_0 \\ x \in \Omega}} \int_{\gamma}
    \frac{\eta(y)}{x-y}\, dy = -\frac{1}{2}\eta(x_0) - 
    \frac{1}{2\pi i} \int_{\gamma} \frac{\eta(y)}{x-y} \, dy,
    \quad x_0 \in \gamma,
    \label{eqn:SP}
\end{align}
where the last integral is interpreted in the principal value sense.
Here we are assuming that $\Omega$ is the bounded region interior to
$\gamma$. Once $v^{-}$ is calculated, $v(x)$ and its derivatives can be
determined by its boundary data alone using the Cauchy Integral Theorem
\begin{subequations}
  \label{eqn:cauchy}
  \begin{alignat}{3}
  \label{eqn:cauchyv}
  v(x) &= \frac{1}{2\pi i}\int_{\gamma} 
    \frac{v^{-}(y)}{y-x} \,dy, \\
  v'(x) &= \frac{1}{2\pi i} \int_{\gamma}
    \frac{v^{-}(y)}{(y-x)^2} \, dy, \\
  v''(x) &= \frac{1}{\pi i} \int_{\gamma}
    \frac{v^{-}(y)}{(y-x)^3} \, dy.
  \end{alignat}
\end{subequations}

@@

The derivatives of the Cauchy integral in~\eqref{eqn:cauchy} can be
approximated with a uniformly spectral error using similar quadrature
rules. For $x \in \Omega$, 
\begin{align}
  v'(x) &\approx \left(\frac{1}{2\pi i}\sum_{j=1}^{N}
    \frac{v^{-}(y_j) - v(x)}{(y_j-x)^2} w_j \right)
  \Bigg/
  \left(\frac{1}{2\pi i}\sum_{j=1}^{N} \frac{1}{y_j-x} w_j\right), 
  \label{eqn:D1vBary} \\
  v''(x) &\approx \left(\frac{2}{2\pi i}\sum_{j=1}^N 
    \frac{v^{-}_{j} - v(x) - (y_j-x)v'(x)}{(y_j-x)^3}w_j \right)
    \Bigg/
    \left(\frac{1}{2\pi i}\sum_{j=1}^N \frac{1}{y_j-x}w_j\right).
  \label{eqn:D2vBary}
\end{align}
We note that equations~\eqref{eqn:vBary},~\eqref{eqn:D1vBary},
and~\eqref{eqn:D2vBary} all assume that $x \in \Omega$, where $\Omega$
is the bounded region interior to the Jordan curve $\gamma$. However, in
our application, when $\gamma$ is one of the eroding bodies, $x$ is in
the exterior region of the Jordan curve. In this case, slightly
different identities are used, but they all guarantee that the trapezoid
rule achieves spectral accuracy with an error that is independent of
$x$. A complete description of the quadrature rules
for~\eqref{eqn:vBary} and~\eqref{eqn:D1vBary} are described by Barnett,
Wu, and Veerapaneni~\cite{bar-wu-vee2015}, and the quadrature rule
for~\eqref{eqn:D2vBary} is described in our previous
work~\cite{chiu2020viscous}.

@@

Like the permeability, the tortuosity provides a macroscopic perspective
of the porous medium. We define the tortuosity to be the average length
of streamlines passing through the region $[-1,1] \times [-1,1]$. In
particular, for a particular eroded geometry with velocity field $\uu$,
we calculate streamlines $\pp(t)$ that satisfy
\begin{align}
  \dot{\pp}(t) = \uu(\pp), \quad \pp(0) = (-1,y_0),
\end{align}
where $y_0 \in (-1,1)$. The streamlines are calculated with a fourth-order Runge-Kutta method. Then, the length of each streamline, which only depends on its initial $y$-coordinate, is
\begin{align}
  \lambda(y_0) = \int_{0}^{t(y_0)} \|\pp(t)\| \,dt.
\end{align}
The integral limit $t(y_0)$ is chosen so that the $x$-coordinate of
$\pp(t(y_0))$ is $+1$. Then, the longitudinal tortuosity is
\begin{align}
  T_1 = \frac{1}{2}\left(\int_{S}u(-1,y)\lambda(y)\,dy \right)
  \Bigg/
  \left(\int_{S}u(x_0,y)\,dy \right).
  \label{eqn:tortuosity_x}
\end{align} 
\todo[inline]{What is $x_0$ in the denominator? Should that be -1?}
The factor of $1/2$ guarantees that $T_1 \geq 1$, and $T_1 = 1$ if and only if no grains are present. Identical to how we calculate the transverse permeability, we rotate \nick{the configuration of} eroded bodies by 90 degrees to compute a transverse tortuosity $T_2$.

@@

With the numerical methods in place, we now present results on how fluid-mechanical erosion alters porous-media properties over time. We will first discuss the results from a single simulation and then generalize to statistical results from ensembles of simulations.

@@

To begin, we discuss a single simulation of 80 bodies eroding in Stokes flow. For reference, the simulation discussed here is the same one shown in Fig.~\ref{fig1}. The initial configuration consists of 80 circular bodies having randomized sizes and positions. The material-removal process alters both the shape of individual bodies as well as the overall structure of the pore network transmitting the flow, as can be seen in Fig.~\ref{fig1}. Initially, the solid bodies occupy $60\%$ of the area, or the porosity is $\varphi = 0.4$, corresponding to a relatively dense packing; see Fig.~\ref{fig1} for a visual. Figure~\ref{fig2}(a) shows how the solid-body area decreases, or equivalently the porosity increases, over time as the bodies erode. The increasing porosity serves as a convenient proxy for dimensionless time that is insensitive to the end conditions setting the strength of the flow (i.e.~whether we specify the pressure drop or the end velocity to be constant in time). We will henceforth use porosity, $\varphi$, on the horizontal axis of many plots to represent increasing time.

%---------------------------------------------------------%
\begin{figure*}%[htbp]
\centering
\includegraphics[width = 0.99 \textwidth]{./figs/fig2.pdf}
\caption{A single simulation of porous-medium erosion. The initial configuration consists of 80 circular bodies of random size and position, as seen in Fig.~\ref{fig1}. a) As the medium erodes, the fraction of solid-body area decreases, or, equivalently, the porosity increases. b) Resistance to flow can be characterized by either the resistivity (solid) or the cumulative drag (dots), the two of which are related through \eqref{perm_drag}. Both are seen to decrease as the medium erodes, and relationship \eqref{perm_drag} is confirmed by the simulation data. The figure also shows the resistivity in transverse direction (dashed).
 \label{fig2}
}
\end{figure*}
%---------------------------------------------------------%
% Data from 80-4

As the bodies disintegrate and give way to wider pores, the resistance to flow decreases as seen in Fig.~\ref{fig2}(b). The resistance to flow can be quantified in two separate ways, as outlined in Section \ref{sec:medium}, namely by computing the total drag or by extracting the permeability. As discussed in Section \ref{sec:medium}, the drag is a microscopic quantity that can be computed with high accuracy and without ambiguity in our Stokes-based simulations. The permeability, on the other hand, relies on a few approximations, but the extraction of such bulk quantities will be more valuable in characterizing the evolving porous medium. As such, it is particularly useful to compare these two perspectives so that the assumptions underlying the coarse-graining process can be assessed. In particular, we aim to test the approximate formula \eqref{perm_drag}, relating the longitudinal drag $F_1$ to the inverse permeability $1/k_{11}$, also known as the {\em resistivity}. Figure \ref{fig2}(b) indeed shows close agreement between these two quantities over the entire duration of the simulation, thus confirming the ability to extract medium properties during the erosion process.

In Fig.~\ref{fig2}(b) we also show the transverse resistivity, $1/k_{22}$, as it decreases over the course of the simulation. Notice that this transverse resistivity exceeds the longitudinal resistivity by a significant margin, indicating that the configuration provides greater resistance to flow in the vertical direction. This trend fits the intuition that horizontally-aligned channels transmit flow more easily in the longitudinal direction.

%---------------------------------------------------------%
\begin{figure*}%[htbp]
\centering
\includegraphics[width = 0.99 \textwidth]{./figs/fig3.pdf}
\caption{
The permeability and anisotropy of the same 80-body simulation featured in Figs.~\ref{fig1} and \ref{fig2}. a) Both the longitudinal (solid) and transverse (dashed) permeability increase over time as the bodies erode. Also shown in the figure is the longitudinal and transverse permeabilities of the corresponding configurations of circles, which allows us to extract the configurational component of anisotropy. b) The anisotropy of the medium increases as the bodies erode and preferentially allow flow in the longitudinal direction. The total anisotropy is the sum of configurational and shape components. The shape contribution is larger throughout most of the simulation, but near the end of the simulation, the configurational contribution grows large as the channel-structure dominates.
\label{fig3}
}
\end{figure*}
%---------------------------------------------------------%
% Data from 80-4
%c) The relationship between permeability $k$ and porosity $\phi$ for the eroding medium. Both quantities increase with time as the medium erodes, but at different rates due to changes in the microstructure and channelization of the medium.

Figure \ref{fig3}(a) shows more directly how permeability in the longitudinal, $k_{11}$ and transverse, $k_{22}$ directions increase over time as the medium erodes. Once again, the relationship $k_{11} > k_{22}$ indicates a higher propensity for flow in the longitudinal direction. Notice that the permeability increases by nearly 5 orders of magnitude over the course of the simulation, indicating that erosion substantially alters medium properties.

As given in \eqref{eq:anis}, the ratio of longitudinal to transverse permeability defines the anisotropy of the medium, $\anis = k_{11} / k_{22}$, which is plotted in Fig.~\ref{fig3}(b) over the course fo the simulation (solid curve). Early in the simulation the anisotropy is nearly one, as the initial configuration of randomly placed circles has no preferred flow direction. The anisotropy then increases as heterogeneous rates of erosion promote longitudinal flow over transverse flow. The anisotropy peaks at a value of $\anis \approx 4$ later in the simulation. At very late times, the anisotropy approaches unity again as the bodies completely vanish and return the system to a state of no preferred flow direction.

>>>>

\author[TAMU]{Shang-Huan Chiu}
%\affiliation{Department of Scientific Computing, Florida State University, Tallahassee, FL, 32306.}

@@

\address[Colgate]{Colgate University}
\address[FSU]{Florida State University}
\address[TAMU]{Texas A\&M-San Antonio}

@@

Flow-induced erosion acts across a range of scales in the natural world, from massive geological structures sculpted by wind or water~\cite{abrams2009growth, perkins2015amplification, mac2020ultra, sharma2022alcove, mac2022morphological}, to mesoscopic patterns formed by surface or internal flows~\cite{berhanu2012shape, bertagni2021hydrodynamic, weady2022anomalous}, and down to granular and porous networks slowly disintegrating in groundwater flows~\cite{chiu2020viscous, szymczak2009wormhole, jager2017channelization, grodzki2019reactive, bizmark2020multiscale, derr2020flow, zareei2022temporal}. The associated nonlinear feedback between changing shapes and the surrounding flows can imprint across all of these scales, affecting large-scale features as well as small-scale ones, such as the microstructure of porous materials. Porous media encountered in nature, for example, typically exhibit material {\em anisotropy} in that they permit seepage flow in certain directions more easily than in others. Typical materials are more permeable to flow in the longitudinal (or horizontal) direction than transversely by a factor of 5--20~\cite{bear1988dynamics, anderson2015applied}. Most commonly, this material anisotropy is attributed to the sedimentation process, in which, due to the fluid-structure interaction, non-spherical particles tend to settle with their long axis parallel to the plan of deposition~\cite{bear1988dynamics}. Controlled experiments, however, have not been performed to test this hypothesis, and other mechanisms may be at work. Here, we use highly-accurate numerical simulations to examine an alternative, and possibly complementary, mechanism: namely, that the flow-induced erosion of the medium's solid constituents contributes to its overall anisotropy.

@@

The outline of the paper is as follows. In Section~\ref{sec:formulation} we discuss the governing equations for the fluid flow and for interface evolution. In Section~\ref{sec:DLP} we discuss the numerical methods. In Section~\ref{sec:medium} we measure porous-medium properties, in particular permeability. In Section~\ref{sec:results} we provide results and we conclude in Section~\ref{sec:conclusion}.

@@

%-----------------------------------------------------------------------------------------------%
\section{Boundary Integral Equation and Cauchy Integral Formulation}
\label{sec:DLP}
%-----------------------------------------------------------------------------------------------%
To accurately and efficiently solve the Stokes equations~\eqref{eqn:StokesEq}, we reformulate the system as a boundary integral equation (BIE). A BIE formulation has the advantage that all the unknowns are on the one-dimensional boundaries of the domain. That is, only the boundary of the complex geometry must be discretized which we do with a spectrally accurate Fourier basis. Applying the same approach as our previous works~\cite{chiu2020viscous, quaife2018boundary}, we represent the velocity as the sum of a double-layer potential and a combination of Stokeslets and rotlets~\cite{pow-mir1987}
\begin{align}
  \uu(\xx) = \DDD[\eeta](\xx) + \sum_{\ell=1}^{M}\left(
    S[\llambda_\ell](\xx) + R[\xi_\ell](\xx)\right), 
    \quad \xx \in \Omega,
  \label{eqn:velocity}
\end{align}
where 
\begin{align}
  \DDD[\eeta](\xx) = \frac{1}{\pi}\int_{\bd\Omega} 
  \frac{\rr \cdot \nn}{\rho^2} \frac{\rr \otimes \rr}{\rho^2}
  \eeta(\yy) \, ds_\yy,
  \label{eqn:velocityDLP}
\end{align}
where $\rr = \xx - \yy$ and $\rho = \|\rr\|$. Note that $\bd\Omega$ includes both the eroding bodies and the outer boundary. The Stokeslets and the rotlets are
\begin{align}
  S[\lambda_\ell](\xx) = \frac{1}{4\pi} \left(-\log \rho_\ell +
  \frac{\rr_\ell \otimes \rr_\ell}{\rho_\ell^2} \right) \llambda_\ell,
  \quad
  R[\xi_\ell](\xx) = \frac{\rr_\ell^\perp}{\rho_\ell^2}\xi_\ell,
\end{align}
respectively, where $\rr_\ell = \xx - \cc_\ell$, $\rho_\ell = \|\rr_\ell\|$, and $\cc_\ell$ is a point inside body $\ell$. If the density function, Stokeslets, and rotlets satisfy the second-kind boundary integral equation
\begin{subequations}
  \label{eqn:FredBIE}
  \begin{alignat}{3}
    \UU(\xx) &= -\frac{1}{2}\eeta(\xx) + \DDD[\eeta](\xx) + 
      \sum_{\ell=1}^{M}\left(S[\llambda_\ell](\xx) + R[\xi_\ell](\xx)\right), 
      \quad &&\xx \in \Gamma, \\
    \mathbf{0} &= -\frac{1}{2}\eeta(\xx) + \DDD[\eeta](\xx) + 
      \sum_{\ell=1}^{M}\left(S[\llambda_\ell](\xx) + R[\xi_\ell](\xx)\right), 
      &&\xx \in \gamma_\ell, \: \ell = 1,\ldots,M, \\
    \llambda_\ell &= \int_{\gamma_\ell} \eeta(\yy) \, ds_\yy, 
      &&\ell = 1,\ldots,M, \\
    \xi_\ell &= \int_{\gamma_\ell} (\yy - \cc_\ell)^\perp \cdot 
      \eeta(\yy) \, ds_\yy, &&\ell = 1,\ldots,M,
  \end{alignat}
\end{subequations}
then the representation~\eqref{eqn:velocity} satisfies the Stokes equations with the required boundary conditions~\eqref{eqn:StokesEq}. We solve~\eqref{eqn:FredBIE} by discretizing $\gamma_\ell$ and $\Gamma$ at equispaced collocation points, and then replacing the integrals with quadrature rules. This results in a linear system with a mesh-independent condition number, and it is solved iteratively with GMRES.

@@

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\subsection{Quadrature for Cauchy integrals}
The accuracy of our method is determined by the accuracy of the
quadrature rule applied to equation~\eqref{eqn:FredBIE}. Since we have
written the Stokes double-layer potential
velocity~\eqref{eqn:velocityCauchy}, deformation
tensor~\eqref{eqn:deformationCauchy}, and
vorticity~\eqref{eqn:vorticityCauchy} as a sum of Cauchy integrals and
their derivatives, the accuracy of the method depends on the accuracy of
computing a general Cauchy integral
\begin{align}
  v(x) = \frac{1}{2\pi i} \int_{\gamma} \frac{\eta(y)}{x-y} \, dy.
  \label{eqn:cauchy1}
\end{align}
Here we describe a quadrature formulae that was first used to
approximate analytic functions~\cite{ioa-pap-per1991}, and then extended
to Stokes layer potentials~\cite{bar-wu-vee2015}. The quadrature method
requires the boundary data of the analytic function~\eqref{eqn:cauchy1}
which satisfies the Sokhotski-Plemelj jump relation
\begin{align}
  v^{-}(x_0) = \lim_{\substack{x \rightarrow x_0 \\ x \in \Omega}} \int_{\gamma}
    \frac{\eta(y)}{x-y}\, dy = -\frac{1}{2}\eta(x_0) - 
    \frac{1}{2\pi i} \int_{\gamma} \frac{\eta(y)}{x_0-y} \, dy,
    \quad x_0 \in \gamma,
    \label{eqn:SP}
\end{align}
where the last integral is interpreted in the principal value sense.
Here we are assuming that $\Omega$ is the bounded region interior to
$\gamma$. Once $v^{-}$ is calculated, $v(x)$ and its derivatives can be
determined by its boundary data alone using the Cauchy Integral Theorem
\begin{subequations}
  \label{eqn:cauchy}
  \begin{alignat}{3}
  \label{eqn:cauchyv}
  v(x) &= \frac{1}{2\pi i}\int_{\gamma} 
    \frac{v^{-}(y)}{y-x} \,dy, \\
  v'(x) &= \frac{1}{2\pi i} \int_{\gamma}
    \frac{v^{-}(y)}{(y-x)^2} \, dy, \\
  v''(x) &= \frac{1}{\pi i} \int_{\gamma}
    \frac{v^{-}(y)}{(y-x)^3} \, dy.
  \end{alignat}
\end{subequations}

@@

The derivatives of the Cauchy integral in~\eqref{eqn:cauchy} can be
approximated with a uniformly spectral error using similar quadrature
rules. To summarize, for $x \in \Omega$, 
\begin{align}
  v'(x) &\approx \left(\frac{1}{2\pi i}\sum_{j=1}^{N}
    \frac{v^{-}(y_j) - v(x)}{(y_j-x)^2} w_j \right)
  \Bigg/
  \left(\frac{1}{2\pi i}\sum_{j=1}^{N} \frac{1}{y_j-x} w_j\right), 
  \label{eqn:D1vBary} \\
  v''(x) &\approx \left(\frac{2}{2\pi i}\sum_{j=1}^N 
    \frac{v^{-}_{j} - v(x) - (y_j-x)v'(x)}{(y_j-x)^3}w_j \right)
    \Bigg/
    \left(\frac{1}{2\pi i}\sum_{j=1}^N \frac{1}{y_j-x}w_j\right),
  \label{eqn:D2vBary}
\end{align}
We note that equations~\eqref{eqn:vBary},~\eqref{eqn:D1vBary},
and~\eqref{eqn:D2vBary} all assume that $x \in \Omega$, where $\Omega$
is the bounded region interior to the Jordan curve $\gamma$. However, in
our application, when $\gamma$ is one of the eroding bodies, $x$ is in
the exterior region of the Jordan curve. In this case, slightly
different identities are used, but they all guarantee that the trapezoid
rule achieves spectral accuracy with an error that is independent of
$x$. A complete description of the quadrature rules
for~\eqref{eqn:vBary} and~\eqref{eqn:D1vBary} are described by Barnett,
Wu, and Veerapaneni~\cite{bar-wu-vee2015}, and the quadrature rule
for~\eqref{eqn:D2vBary} is described in our previous
work~\cite{chiu2020viscous}.

@@

Like the permeability, the tortuosity provides a macroscopic perspective
of the porous medium \cite{souzy2020velocity}. We define the tortuosity to be the average length
of streamlines passing through the region $[-1,1] \times [-1,1]$. In
particular, for a particular eroded geometry with velocity field $\uu$,
we calculate streamlines $\pp(t)$ that satisfy
\begin{align}
  \dot{\pp}(t) = \uu(\pp), \quad \pp(0) = (-1,y_0),
\end{align}
where $y_0 \in (-1,1)$. The streamlines are calculated with a fourth-order Runge-Kutta method. Then, the length of each streamline, which only depends on its initial $y$-coordinate, is
\begin{align}
  \lambda(y_0) = \int_{0}^{t(y_0)} \|\pp(t)\| \,dt.
\end{align}
The integral limit $t(y_0)$ is chosen so that the $x$-coordinate of
$\pp(t(y_0))$ is $+1$. Then, the longitudinal tortuosity is
\begin{align}
  T_1 = \frac{1}{2}\left(\int_{S}u(-1,y)\lambda(y)\,dy \right)
  \Bigg/
  \left(\int_{S}u(-1,y)\,dy \right).
  \label{eqn:tortuosity_x}
\end{align} 
The factor of $1/2$ guarantees that $T_1 \geq 1$, and $T_1 = 1$ if and only if no grains are present. Identical to how we calculate the transverse permeability, we rotate the configuration of eroded bodies by 90 degrees to compute a transverse tortuosity $T_2$.

@@

With the numerical methods in place, we now present results on how fluid-mechanical erosion alters porous-media properties over time. We first discuss the results from a single simulation and then generalize to statistical results from ensembles of simulations.

@@

To begin, we discuss a single simulation of 80 bodies eroding in Stokes flow. The simulation discussed here is the same one shown in Figure~\ref{fig1}. The initial configuration consists of 80 circular bodies having randomized sizes and positions. The material-removal process alters both the shape of individual bodies as well as the overall structure of the pore network transmitting the flow, as can be seen in Figure~\ref{fig1}. Initially, the solid bodies occupy $60\%$ of the area, or the porosity is $\varphi = 0.4$, corresponding to a relatively dense packing; see Figure~\ref{fig1} for a visual. Figure~\ref{fig2}(a) shows how the solid-body area decreases, or equivalently the porosity increases, over time as the bodies erode. The increasing porosity serves as a convenient proxy for dimensionless time that is insensitive to the end conditions setting the strength of the flow (i.e.~whether we specify the pressure drop or the end velocity to be constant in time). We will henceforth use porosity, $\varphi$, on the horizontal axis of many plots to represent increasing time.

%---------------------------------------------------------%
\begin{figure*}%[htbp]
\centering
\includegraphics[width = 0.99 \textwidth]{./figs/fig2.pdf}
\caption{A single simulation of porous-medium erosion. The initial configuration consists of 80 circular bodies of random size and position, as seen in Figure~\ref{fig1}. a) As the medium erodes, the fraction of solid-body area decreases, or, equivalently, the porosity increases. b) Resistance to flow can be characterized by either the resistivity (solid) or the cumulative drag (dots), the two of which are related through~\eqref{perm_drag}. Both are seen to decrease as the medium erodes, and relationship~\eqref{perm_drag} is confirmed by the simulation data. The figure also shows the resistivity in transverse direction (dashed).
\label{fig2}
}
\end{figure*}
%---------------------------------------------------------%
% Data from 80-4

As the bodies disintegrate and give way to wider pores, the resistance to flow decreases as seen in Figure~\ref{fig2}(b). The resistance to flow can be quantified in two separate ways, as outlined in Section~\ref{sec:medium}, namely by computing the total drag or by extracting the permeability. As discussed in Section~\ref{sec:medium}, the drag is a microscopic quantity that can be computed with high accuracy and without ambiguity in our Stokes-based simulations. The permeability, on the other hand, relies on a few approximations, but the extraction of such bulk quantities will be more valuable in characterizing the evolving porous medium. As such, it is particularly useful to compare these two perspectives so that the assumptions underlying the coarse-graining process can be assessed. In particular, we aim to test the approximate formula~\eqref{perm_drag}, relating the longitudinal drag $F_1$ to the inverse permeability $1/k_{11}$, also known as the {\em resistivity}. Figure~\ref{fig2}(b) indeed shows close agreement between these two quantities over the entire duration of the simulation, thus confirming the ability to extract medium properties during the erosion process.

In Figure~\ref{fig2}(b) we also show the transverse resistivity, $1/k_{22}$, as it decreases over the course of the simulation. Notice that this transverse resistivity exceeds the longitudinal resistivity by a significant margin, indicating that the configuration provides greater resistance to flow in the vertical direction. This trend fits the intuition that horizontally-aligned channels transmit flow more easily in the longitudinal direction.

%---------------------------------------------------------%
\begin{figure*}%[htbp]
\centering
\includegraphics[width = 0.99 \textwidth]{./figs/fig3.pdf}
\caption{
The permeability and anisotropy of the same 80-body simulation featured in Figures~\ref{fig1} and~\ref{fig2}. a) Both the longitudinal (solid) and transverse (dashed) permeability increase over time as the bodies erode. Also shown is the longitudinal and transverse permeabilities of the corresponding configurations of circles, which allows us to extract the configurational component of anisotropy. b) The anisotropy of the medium increases as the bodies erode and preferentially allow flow in the longitudinal direction. The total anisotropy is the sum \todo[inline]{do you mean product} of configurational and shape components. The shape contribution is larger throughout most of the simulation, but near the end of the simulation, the configurational contribution grows large as the channel-structure dominates.
\label{fig3}
}
\end{figure*}
%---------------------------------------------------------%
% Data from 80-4
%c) The relationship between permeability $k$ and porosity $\phi$ for the eroding medium. Both quantities increase with time as the medium erodes, but at different rates due to changes in the microstructure and channelization of the medium.

Figure~\ref{fig3}(a) shows more directly how permeability in the longitudinal, $k_{11}$ and transverse, $k_{22}$ directions increase over time as the medium erodes. Once again, the relationship $k_{11} > k_{22}$ indicates a higher propensity for flow in the longitudinal direction. Notice that the permeability increases by nearly five orders of magnitude over the course of the simulation, indicating that erosion substantially alters medium properties.

As given in~\eqref{eq:anis}, the ratio of longitudinal to transverse permeability defines the anisotropy of the medium, $\anis = k_{11} / k_{22}$, which is plotted in Figure~\ref{fig3}(b) over the course of the simulation (solid curve). Early in the simulation the anisotropy is nearly one, as the initial configuration of randomly placed circles has no preferred flow direction. The anisotropy then increases as heterogeneous rates of erosion promote longitudinal flow over transverse flow. The anisotropy peaks at a value of $\anis \approx 4$ later in the simulation. At very late times, the anisotropy approaches unity again as the bodies completely vanish and return the system to a state of no preferred flow direction.

As discussed in Section~\ref{subsec:anis}, the anisotropy of the medium results from the combination of {\em shape} anisotropy, at the individual grain level, and {\em configurational anisotropy}, at the large scale. As detailed in Section~\ref{subsec:anis}, the configurational anisotropy can be extracted through a process in which, at any point in time, the configuration of partially-eroded bodies is replaced with a collection of circles having the same areas and centers of mass. We perform this process and show in Figure~\ref{fig3}(a) the permeabilities, $k_{11}^{(c)}$ and $k_{22}^{(c)}$, of the corresponding circle configurations. As seen in the figure, the permeabilities of the circle configurations always lie in between the values, $k_{11}$ and $k_{22}$, of the true, eroded medium. The {\em configurational} anisotropy is then computed as the permeability ratio of the circle configuration, $\anis_C = k_{11}^{(c)} / k_{22}^{(c)}$, and then the {\em shape} anisotropy can be determined through $\anis_S = \anis/\anis_C$. As seen in Figure~\ref{fig3}(b), the shape anisotropy is the larger factor throughout the majority of the simulation, implying that the shape of individual grains plays a larger role than their relative positions. It is only during the last 10\% of the simulation that the configurational anisotropy grows significantly and surpasses the shape anisotropy. This relative growth of the configurational anisotropy corresponds to the strong channelization seen in the last panel of Figure~\ref{fig1}.

The late emergence of substantial configurational anisotropy may seem
like a spurious effect that is particular to the one simulation shown in
Figure~\ref{fig3}. However, statistical analysis of many simulations will show this feature to emerge robustly in simulations featuring a relatively large number of bodies $M \ge 60$.

%---------------------------------------------------------%
\begin{figure*}%[htbp]
\centering
\includegraphics[width = 0.99 \textwidth]{./figs/fig4.pdf}
\caption{
Tortuosity measurements of the same 80-body simulation featured in Figures~\ref{fig1} and~\ref{fig2}. a) As the bodies erode, the transverse tortuosity $T_2$ exceeds the longitudinal component, indicating more convoluted paths for tracers traveling vertically, transverse to the flow. Also shown are the measurements for the configurations of circles. b) The ratio $(T_2-1)/(T_1-1)$ serves as a counterpart to the anisotropy. Like anisotropy, this ratio grows large over the course of erosion. It can be decomposed into configurational and shape components in a similar manner.
\label{fig4}
}
\end{figure*}
%---------------------------------------------------------%
% Data from 80-4
